---
title: "assignment_5_LNM"
output: html_document
date: "2023-02-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load cancer data from cancer.csv and define a new column dead that is either 1 for dead and 0 for not. 

```{r}
cancer_data <- read.csv("~/Downloads/cancer.csv", stringsAsFactors=TRUE)[,-1]
cancer_data$death <- ifelse(cancer_data$status==2,1,0)
head(cancer_data)
#status = 2=death occured 
#status=1=no event (they are censored)
```

### A. Fit Kaplan-Meier method on time and dead columns and plot survival function.

```{r}
library(survival)

km.model<-survfit(Surv(time, death) ~ 1,data=cancer_data,type="kaplan-meier")
km.model
plot(km.model, conf.int=F, xlab='Months',ylab="%alive = s(t)", main="KM-Model-LNM")
```

### B. Median Survival time is 310 months. 
```{r}
km.model
```
### C. Print Kaplan-Meier estimates (timeline vs k-m estimate)

```{r}
time_df <- as.data.frame(cbind(km.model$time,km.model$surv))
colnames(time_df) <- c('Time','K-M estimate')
time_df
```
### D. Create a combined plot where male and female groups’ Kaplan-Meier survival functions displayed

```{r warning=FALSE}
km.model2<-survfit(Surv(time,death) ~ sex,data=cancer_data, type="kaplan-meier")


{plot(km.model2, conf.int=F, xlab='Months',ylab="%alive = s(t)", main="KM-Model-LNM", las=1, lwd=2, col=c('green','blue'),mark.time=TRUE)
legend(400,0.95,legend=c('Male','Female'),lty=1, lwd=2, col=c('green','blue'),bty="", cex=0.6)}
```

### E. Conduct a log-rank test and test the null hypothesis that there is no difference in survival between 2 groups, namely male vs female. We reject the null hypothesis because the p value is less than 0.05 -- there is a difference in survival between the two groups. 

```{r}

survdiff(Surv(time, death) ~ sex,data=cancer_data)
```

